---
title: "Caching issues for algoithms with random parameters in R"
author: "Michael Lai"
date: "2023-10-15"
output: 
  pdf_document: 
    number_sections: yes
---

```{r setup, include=F}
knitr::opts_chunk$set(cache = F, echo = T)
```

This is an experiment to show how caching behaves in chunks when a random seed is used. Each trial is a function to generate 6 random numbers of mean 100 and standard deviation 15, with various results according to different settings. 

Note because the caching option is set at each chunk level, the global setting no longer matters.

The first trial is just a computation without caching nor seed manipulation. It should produce different numbers every time it is knitted.

Scenario 1 - Chunk option: cache=F
```{r not_cached, cache=F}
rnorm(6,100,15)
```

The second trial is the same computation but with caching instead. After the first attempt, subsequent runs should produce the same numbers because it just retrieves the data in the cache rather than rerunning the code.

Scenario 2 - Chunk option: cache=T
```{r cached_no_set_seed, cache=T}
rnorm(6,100,15)
```

The thrid trial is the same computation without caching. Surprisingly, despite not using cache the results are the same as the first run.

Scenario 3 - Chunk option: cache=F
```{r not_cached_no_set_seed, cache=F}
rnorm(6,100,15)
```

Further investigation shows this is because the cache in the scenario 2 also keeps the  seed, which is then used to generate this pseudo-random number set. Hence the newly generated set is still the same as before.

```{r random_seed, cache=F}
.Random.seed[1:6]
```

In order to generate a new random set of numbers, apart from setting the caching option to false, the seed is also reset to null to clear the influence from the previous chunk. This is confirmed by the new seed set below. 

```{r new_random_seed, cache=F}
set.seed(NULL)
.Random.seed[1:6]
```

Then the regenerated number set will be different each time the document is knitted.

Scenario 4 - Chunk option: cache=F
```{r not_cached_reset_seed, cache=F}
rnorm(6,100,15)
```

Finally, if the chunk is cached it does not matter whether the seed is reset. The seed and the number set will still be the same as before because the code is not rerun.

Scenario 5 - Chunk option: cache=T
```{r cached_reset_seed, cache=T}
set.seed(NULL)
.Random.seed[1:10]
rnorm(6,100,15)
```

Also it is noted that the cache used in knitting is not the same as the one used in the preview of the markdown notebook (those results shown in snippets). They will not be the same if there is any randomness involved in the algorithm. If the exact output has to be ensured, a seed has to be set at the beginning of each chunk.